---
title: "Country Analysis by GDP and Income Group for 2012"
author: "James Hosker"
date: "March 22, 2017"
output: 
  pdf_document:
  keep_md: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
This report examines data from the World Bank in 2012 on Gross Domestic Product (GDP) combined with country specific Federal Statistics by country.   A summary of our methods and analyses are below:

I.   We successfully merge and clean two datasets from the World Bank:  2012 GDP data of 190 countries and 2012 Federal Statistics of 235 countries.   This resulting dataset for GDP vs. Federal Statistics provides data on 190 countries spanning several income groups for low, middle, upper and high income countries.  For more information on the raw datasets, see the References section at the end of this report.

II.  We rank the 2012 cleaned dataset by increasing GDP in USD millions and break it out into income group categories for each country:  low income; lower middle income; upper middle income; high income: nonOECD; and high income: OECD.  OECD is the classification from the Organization for Economic Cooperation and Development (OECD), for more information see
[\textcolor{blue}{OECD website link}](http://www.oecd.org/about/membersandpartners/list-oecd-member-countries.htm]).

>1. In merging the datasets, we provide how many country code IDs matched.

>2. We show as an example the 13th country in GDP rank after sorting each country by GDP in USD millions in ascending order.

>3. For the high income: OECD and nonOECD country income groups, we compared the mean and summary statistics for GDP rankings and GDP in USD millions.  

>4. We plot GDP for many countries and plot GDPs by country income group.  In addition, we provide summary statistics GDP ranks by country income groups.

>5.  We separate the country GDP ranks into 5 separate quantile groups, making a table of increasing GDPs with country income groups.   In addition, we determine how many countries are lower middle income but among the 38 nations with highest GDPs globally.  Perhaps these countries will have income group upgrades in the future, if GDP continues to remain high and expand.  

III. Finally, we provide our conclusions and recommendations for future work.

# Folder Description

The URL with the code and our analysis is on the GitHub website:  [\textcolor{blue}{https://github.com/jjhosker/CaseStudy1}](https://github.com/jjhosker/CaseStudy1.git).  It contains the following important files:

>1.  README:  his file provides instructions on how to run the R code and provides information in detail of each of the files in the GitHub repository.

>2.  RMakefile_CaseStudy1.R:  This file can be found in the Analysis sub-directory under this GitHub repository.   It is the main R makefile that takes the cleaned data and performs all the analysis required for this project and in this report. In cleaning the data, we read in the csv files and in some cases re-name variables and reset data types.  In addition, we remove NAs and blanks in the data.  Finally, we merge the data by the unique country code in both datasets and create and output one cleaned and merged dataset (GDPFEDclean).

>3.  Makefile_GatherData.R:  This file can be found in the Analysis/Data sub-directory under this GitHub repository.  This R makefile installs all required R-libraries and then loads the R-libraries.  Second, this R makefile downloads in the csv data files from the World Bank websites using URLs, reads the downloaded csv files, creates the raw data, and creates the cleaned datasets.   Finally, this R makefile merges the GDP with the Federal Statistical dataset and creates one a clean data set.  This makefile then saves all these datasets to the Analysis/Data sub-directory.  

>4.  PlotFunctions_CaseStudy1.R:  In the Analysis subdirectory of this GitHub repository, this R Makefile loads plot functions to plot Figures 1 and 2 used in RMakefile_CaseStudy1.R and CaseStudy1PDF.Rmd.

>5.  There are three data files created below:  two raw datsets downloaded from the World Bank URLs and one cleaned dataset.   In addition, we provide a text file (SessionsInfo.txt) containing all the R session information including the current R version and all the libraries and packages being utilized.  All of the files below can be found in the Analysis/Data sub-directory under this GitHub repository.
>     - Raw:  FEDSTATS_Country.csv
>     - Raw:  GDP_Country.csv
>     - Cleaned:  GDPFEDclean.csv
>     - R version:  SessionInfo.txt

>6.  Finally, the CaseStudy1PDF.Rmd file runs the main RMakefile_CaseStudy1.R and extracts code to generate this "CaseStudy1PDF.pdf"" report.   This report uses code that that is extracted from the main RMakefile_CaseStudy1.R and then run.
>     - To use this Rmd file to create a PDF, the user has to install MikTex (the non-basic version).  The URL with instructions to install MikTex is [\textcolor{blue}{MikText Install website}](https://miktex.org/howto/install-miktex).

# Running the R-Code

Before beginning please review the README.md file in GitHub website [\textcolor{blue}{https://github.com/jjhosker/CaseStudy1}](https://github.com/jjhosker/CaseStudy1).  In short, download and clone the GitHub repository and the sub-directories.  RMakefile_CaseStudy1.R should be executed in the Analysis sub-directory of this GitHub repository; and CaseStudy1PDF.Rmd can be executed from this GitHub repository.  

>1. In the Analysis sub-directory in this repository, execute the main R source code RMakefile_CaseStudy1.R to load all the libraries, read the raw data, create the cleaned data and perform the analysis for this client study.  The csv files will be created in the same directory as the project that you created on your computer.  In addition, we save the R session information and all the libraries and packages being utilized.  
>     - The RMakefile_CaseStudy.R executes the makefile Makefile_GatherData.R, which installs the packages and load the libraries.  A message indicating that the libraries loaded may appear after executing the R code that follows.  If a failure occurs, please check the function "packagelibrary.check" in the Makefile_GatherData.R and/or the SessionInfo.txt file, both found in the Analysis/Data sub-directory.

>2. To generate this report, execute CaseStudy1PDF.Rmd in this GitHub repository.  Please note the required installations of software to generate the PDF is detailed in the prior section, Folder Description. 
 
```{r}
###################################################
## Execute Main RMakefile_CaseStudy1.R Makefile
###################################################
source("Analysis/RMakefile_CaseStudy1.R")
```

# Five Key or Major Points/Questions:    

## Key Point #1:  Country Code IDs that Match & Variables Names
The R code below prints the necessary information around the IDs that match and the variables that we are using.  For more details, see R-code section Key Point/Question #1 in RMakefile_CaseStudy1.R.
```{r}
print(paste("Number of country code IDs that match: ",
 sum(!is.na(GDPFEDclean$GDP_MillUSD)),"."))
print(paste("Number of country code IDs that do not match: ",
 sum(is.na(GDPFEDclean$GDP_MillUSD)),"."))
print(paste("Total Number of Variables: ",length(GDPFEDclean),"."))
cat(paste("Variable names are:", toString(variable.names(GDPFEDclean)[1:4]),"\n",
          "                  ",
          toString(variable.names(GDPFEDclean)[5:7]),"."))
```
>1.  The number of country codes IDs that match between the 2012 World Bank GDP and Federal Statistic datasets is 190.  There are 190 entries in the World Bank GDP dataset and 235 in the Federal Statistic dataset.  Therefore, 45 do not match and are generally small countries or countries that have little GDP, like North Korea (PRK) even though South Korea is part of those countries that do match.   In addition, there are some countries with two entries such as Samoa and American Samoa, in which only Samoa is in both datasets.

>2.  We can see that the out merged dataset that we use 7 major variables for this analysis:  CountryCode, GDPRank, Economy, GDP_MillUSD, Long.Name, Income.Group and Region.

## Key Point #2: GPD & GDP Rank by Country
The R code below finds the 13th lowest ranked country by GDP (ranked lowest to highest).  For more details, see R-code section Key Point #2 in RMakefile_CaseStudy1.R.
```{r}
## find 13th entry into the data frame after sorting
print(paste("13th Ranked Country by GDP is",
 GDPFEDsort$CountryCode[13], ", which is ", GDPFEDsort$Economy[13],"."))
```
We sort the data by ascending GDP from (lowest to highest) and we can find the country with the lowest to highest GDP.  For example, we find that the 13th country from the country with the lowest GDP is St. Kitts and Nevis (KNA) with a 2012 GDP of $767 Million. 

## Key Point #3:  Mean Stats for High Income OECD & non-OECD Countries
The R code below finds the mean GDP rank of two income groups.  For more details, see R-code section Key Point #3 in RMakefile_CaseStudy1.R.
```{r}
cat(paste("The mean GDP rank of the High Income:  OECD income group is",
    round(MeanRankOECD,1),"\n       & the GDP is in", 
    round(MeanGDPOECD,1),"USD Millions"))
cat(paste("The mean GDP rank of the High Income: nonOECD income group is",
    round(MeanRanknonOECD,1),"\n      & the GDP is in", 
    round(MeanGDPnonOECD,1),"USD Millions"))

```
>1.  GDP Rank give highest GDP at 1 and lowest at 190.

>2. Here we analyze the data and find that the mean GDP Rankfor the high income:  OECD countries is 32.97 (higher) and the mean GDP for the high income:  OECD countries is $1,483,917 million.  

>3. Here we analyze the data and find that the mean GDP Rank for the high income:  non-OECD countries is lower at 91.97 (lower) and the mean GDP for the high income:  non-OECD countries is $104,349.8 million. 

\newpage
## Key Point #4:  Graphs & Summary Stats on Country GDPs vs. Income Groups
Here we provide three graphs and one table for GDP and GDP vs. Income Group. GDP in USD Millions is ranked from lowest to highest.  One country gets eliminated since South Sudan (SSD) has an NA for country income group but has a valid GDP of 10,200 USD Million and valid GDP rank.  The number of countries with both GDP and income group reduces to 189 from 190.

This R code below provides a plot for GDP by country income groups.  It also provides a summary table of statistics for each income group by GDP ranks.  For more details, see R-code section Key Point #4 in RMakefile_CaseStudy1.R.  
```{r}
## Plot all GDPs with income group classifications
PlotFig1()
```

>1. Figure 1 is the GDP of all Countries is shown first from Lowest to Highest GDP.  Each country income group is highlighted in a different shape and color.  We also label the countries with the lowest, the 50th, the 100th, the 150th and the four top GDP ranks.  

\newpage
This R code below provides a plot for GDP by country income groups separately.  For more details, see R-code section Key Point #4 in RMakefile_CaseStudy1.R.  
```{r}
## Plot GDP ranks of each income group separately and label lowest to highest by country code
PlotFig2()
```

>2. Figure 2 is the GDP of countries by income group from lowest to highest GDP within each income group in order to get a clearer picture of each income group individually.  In addition, we provide a label for the lowest and highest GDP of each income group by its country code.

\newpage
This R code below provides a plot for GDP by country income groups separately capped at 2,500,000 USD million.  For more details, see R-code section Key Point #4 in RMakefile_CaseStudy1.R.  
```{r}
## Plot GDPs of each income group separately for GDP < $2,500,000 Mill
ggplot(subset(GDPFEDinc, Income.Group %in% c("Low income","Lower middle income",
 "Upper middle income", "High income: nonOECD", "High income: OECD")),
 aes(x=GDPFEDinc$CntryNum, y=GDPFEDinc$GDP_MillUSD, color=Income.Group)) + 
 geom_point() +  ggtitle("Fig.3: GDP by Income Group (GDP < $2,500,000 Mill)") + 
 labs(x="GDP within Each Income Group (Low to High)", y="GDP in USD Millions") +
 ylim(0,2500000)
```

>3. Figure 3 is the GDP of countries by income group from lowest to higher GDP within each income group for GDPs <= 2,500,000 USD million OR 2.5 USD trillion. It shows more clearly the distribution of GDPs for each category capping those GDPs that are displayed (removing 5 countries with very high GDPs). 

This R code below provides summary statistics table of GDP ranks for each country income group.  For more details, see R-code section Key Point #4 in RMakefile_CaseStudy1.R.  
```{r echo=TRUE}
## Create table of GDP ranks by Country Income Group
kable((sumtable), format="pandoc", align='c',row.names = FALSE,col.names = sumclass,
 caption = "Summary Statistics of GDP Ranks by Income Groups", echo = TRUE)
```

>4. In Table 1, we show summary statistics on the GDP ranks of each income group (highest GDP is ranked 1, lowest GDP is ranked 189).  We can see that those classified as low income have the lowest cluster of GDPs (mean GDP rank 133.7).  The lower middle income group (mean GDP rank is 107.7) has a few outliers that are larger than the next higher category of the upper income group.  However, in general, the lower middle income group has GDPs that are near or slightly higher (lower GDP ranks) than the low income group.  The upper middle income (mean GDP rank is 92.13) generally has GDPs higher (lower GDP ranks) than the lower middle income group (excluding the outliers).  In general, the 1st quantiles to 3rd quantile follows a distribution in line with our expectations that lower income countries have higher GDP ranks (lower GDPs) than higher income countries (lower GDP rank, higher GDPs)

>5. Finally, we see that the high income group of OECD countries (mean GDP rank is 32.97) has the higher GDPs of any income group category.  In addition, the high income group with non-OECD countries has a similar GDPs (mean GDP rank 92.17) compared to the upper middle income group (mean GDP rank 92.13).

## Key Point #5: Rank GDP by Quintiles & Compare to Income Groups
This R code provides analysis of ranking GDP by quintiles versus income groups.  For more details, see R code section Key Point #5 in RMakefile_CaseStudy1.R.
```{r echo = TRUE}
## Use kable in knitr to render table of GDP by Quintiles with Income Groups
kable((GDPFEDsort2[,colvars]), format="pandoc", align='c',
 row.names = FALSE, col.names = TableColNames, 
 caption = "Summary Data by Quintile Ranking of GDP ($Mill)", echo = TRUE)  
```

> 1. Table 2 shows the ranking of GDPs from lowest to highest, broken into five Quintiles:  Q1, Q2, Q3, Q4 and Q5, where Q1 is the lowest quintile and Q5 is the highest quintile for GDP.  One exception is that China is still classified as part of the lower middle income group with the second largest GDP in the world.

This R code provides a summary table of the count of each income group for each GDP quintile.  For more details, see R code section Key Point #5 in RMakefile_CaseStudy1.R.
```{r echo=TRUE}
## Count for each quintile the number in each income category
kable((sumquint), format="pandoc", align='c',row.names = FALSE,
 ncol.names = varnameclass, 
caption = "Summary Count of Income Groups in Each Quintile", echo = TRUE)  
```

>2.  In Table 3, we summarize Table 2 by income group counts for each of the quintiles.  In general, GDP rankings match income grouping expectations.  As GDP increases Q1 to Q5, we see more upper middle income countries and more high income countries in the OECD as well as some high income countries that are non-OECD.  Vice-versa as GDP decreases from Q5 to Q1, we see more low income and lower middle income countries.

\newpage
This R code provides a list of the countries that are in the lower middle income group but are one of the top 38 countries by GDP rank.  For more details, see R code section Key Point #5 in RMakefile_CaseStudy1.R.
```{r}
## Find those countries that are in the Lower Middle income but among 
## the top 38 nations with the highest GDP (using GDPRank)
Filtered <- GDPFEDsort2[GDPFEDsort2$GDPRank <= 38 & 
            GDPFEDsort2$Income.Group == "Lower middle income",
            c("CountryCode","Economy","GDPRank")]
## Render Table of Countries meeting the criteria
kable((Filtered), format="pandoc", align='c',row.names = FALSE,
      col.names = c("CountryCode","Economy","GDPRank"), 
      caption = "Countries in Lower Middle Income with GDPRank<=38")  
```

>3.  In Table 4, we find those countries that are in the "Lower Middle income" group that are among the top 38 nations with the highest GDP using the GDP Rank.  These five countries may have the potential to have their income group category upgraded in future years.    

# Conclusions:

In this report, we addressed five major or key points.  

> - First we merge the datasets of GPD with Federal Statistics from the World Bank and find that 190 country codes match.  

> - Second, we rank order each country by GDP and find that the 13th country from the lowest GDP is St. Kitts and Nevis (KNA) with a 2012 GDP of 767 USD Million

> - Third, we find that the mean GDP Rank for the high income:  OECD countries is 32.97 (higher) and the mean GDP is 1,483,917 USD million. In addition, we find that the mean GDP rank for high income:  non-OECD countries is lower at 91.97 (lower) and the mean GDP is 104,349.8 USD million. 

> - Fourth, we find that in general that those classified in the low income group have the lowest GDPs, followed by the lower middle income group with slightly higher GDPs and then followed by the upper middle income group.  Finally, we see that the high income group of OECD countries has the higher GDPs of any income group category.  The high income group with non-OECD countries has a similar GDPs (mean GDP rank 92.17) compared to the upper middle income group (mean GDP rank 92.13).

> - Fifth we show that when we break GDP by rank into five quintiles in ascending order, we generally see that the lower quintiles (Q1) have the lower income and the lower middle income groups, while the upper quintiles (Q5) have the high income group with OECD countries.   One exception is that China is still classified as part of the lower middle income group with the second largest GDP in the world.    
>       - In addition, we found those countries that are in the "Lower Middle income" group and are among the top 38 nations with the highest GDP using the GDP Rank.  These five countries may have the potential to have their income group category upgraded in future years:  Egypt (EGY), Thailand (THA), Indonesia (IDN), India (IND) and China (CHN).

# Future Work
We have explored much in this dataset but there is still more to analyze.  We could analyze GDP or GDP rank to the region the country is in (e.g. Europe & Central Asia, Latin America & Caribbean, East Asia & Pacific, Sub-Saharan Africa, etc.).   In addition, we could look at the last census of each country and compare that to GDP or GDP rank.  Another comparison, is looking at GDP versus the type of government (consolidated or budgetary) or system of trade (general or specific).

# References:
Relevant Data Links/URLs:  We look at two 2012 datasets from the following World Bank websites:

> 1. Country specific GDP Data:  [\textcolor{blue}{World Bank GDP file link}](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv)
>     - For more information on World Bank GDP by country see: [\textcolor{blue}{World Bank GDP website}](http://data.worldbank.org/data-catalog/GDP-ranking-table)  

>2. Country specific Federal Statistical Data:  [\textcolor{blue}{World Bank Federal Stats file link}](https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv)
>     - For more information on World Bank Federal Statistics by country see: [\textcolor{blue}{World Bank Federal Stats website}](http://data.worldbank.org/data-catalog/ed-stats)

